---
title: "Introduction to R mallet"
author: "David Mimno and Måns Magnusson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mallet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
```


## Installation

To use the mallet R package, we need to use `rJava`, an R package for using Java within R (to access the mallet Java code). See details at [github.com/s-u/rJava](https://github.com/s-u/rJava).

Next, install the ```mallet``` R package from CRAN. To install, simply use ```install.packages()```

```{r, eval=FALSE}
install.packages("mallet")
```

## Usage

Depending on the size of your data, it can be so that you need to increase the Java virtual machine (JVM) heap memory to handle larger corpora. To do this, you need to specify how much memory you want to allocate to the JVM using the ```Xmx``` flag. Below is an example of allocating 4 Gb to the JVM. 

```{r, eval=FALSE}
options(java.parameters = "-Xmx4g")
```

To load the package, use ```library()```.

```{r}
library(mallet)
```

### Reading data into R

There are multiple ways to read text data into R. A simple way is to read individual text files into a character vector. Below is an example of reading the different stop list txt files that come with the ```mallet``` package into R as a character vector (that can be used by the ```mallet``` R package as data).

```{r}
# Note this is the path to the folder where the stoplists are stored in the R package.
# Change this path to another directory to read other txt files into R.
directory <- system.file("stoplists", package = "mallet")

files_in_directory <- list.files(directory, full.names = TRUE)

txt_file_content <- character(length(files_in_directory))
for(i in seq_along(files_in_directory)){
  txt_file_content[i] <- paste(readLines(files_in_directory[i]), collapse = "\n")
}
# We can check the content with str()
str(txt_file_content)
```

We will now use the example data set of the State of the Union addresses from 1946 to 2000 that is included with the ```mallet``` R package as a ```data.frame```. This data can be accessed as follows.

```{r}
library(dplyr)
data(sotu)
sotu[["text"]][1:2]
```

Mallet also comes with five different stop list files (see above). We can access the path to these lists as follows.

```{r}
mallet_supported_stoplists()
stopwords_en_file_path <- mallet_stoplist_file_path("en")
```

### Training topic models

As a first step, we need to create an LDA trainer object and supply the trainer with documents. We start by creating a mallet instance list object. 
This function has a few extra options (whether to lowercase or how we define a token). See ```?mallet.import``` for details.

```{r}
sotu.instances <- 
  mallet.import(id.array = row.names(sotu), 
                text.array = sotu[["text"]], 
                stoplist = stopwords_en_file_path,
                token.regexp = "\\p{L}[\\p{L}\\p{P}]+\\p{L}")
```

If the data is already cleaned and we want to use the index of `text.array`, we can supply the `text.array`.

```{r}
sotu.instances.short <- 
  mallet.import(text.array = sotu[["text"]])
```

It is also possible to supply stop words as a character vector. 

```{r}
stop_vector <- readLines(stopwords_en_file_path)
sotu.instances.short <- 
  mallet.import(text.array = sotu[["text"]], 
                stoplist = stop_vector)
```


We first need to create a topic trainer object to fit a model.

```{r}
topic.model <- MalletLDA(num.topics=10, alpha.sum = 1, beta = 0.1)
```

Load our documents. We could also pass in the filename of a saved instance list file we build from the command-line tools.

```{r}
topic.model$loadDocuments(sotu.instances)
```

We use the method `getVocabulary()` to get the model's vocabulary. The vocabulary may be helpful in further curating the stopword list.

```{r}
vocabulary <- topic.model$getVocabulary()
head(vocabulary)
```

Similarly, we can access the word and document frequencies with ```mallet.word.freqs()```.

```{r}
word_freqs <- mallet.word.freqs(topic.model)
head(word_freqs)
```


To optimize hyperparameters (\code{alpha} and \code{beta}) every 20 iterations, after 50 burn-in iterations, we set alpha optimization as follows.

```{r}
topic.model$setAlphaOptimization(20, 50)
```

Now train a model. Note that hyperparameter optimization is on by default. We can specify the number of iterations. Here we'll use a large-ish round number.

```{r}
topic.model$train(200)
```

We can also run through a few iterations where we pick the best topic for each token rather than sampling from the posterior distribution.

```{r}
topic.model$maximize(10)
```

### Analysis of a topic model

To analyze our corpus using our model, we usually want to access the probability of topics per document and the probability of words per topic. By default, these functions return raw word counts. Here we want probabilities, so we normalize and add "smoothing" so that nothing has exactly 0 probability.

```{r}
doc.topics <- mallet.doc.topics(topic.model, smoothed=TRUE, normalized=TRUE)
topic.words <- mallet.topic.words(topic.model, smoothed=TRUE, normalized=TRUE)
```

What are the top words in topic 2? Notice that R indexes from 1 and Java from 0, so this will be the topic that mallet called topic 1.

```{r}
mallet.top.words(topic.model, word.weights = topic.words[2,], num.top.words = 5)
```

Show the largest document with at least 50% tokens belonging to topic 2. Note, since the model is not identified, you might end up with another topic if you run the same code. 

```{r}
docs <- which(doc.topics[,2] > 0.50)
doc_size <- nchar(sotu[["text"]])[docs]
idx <- docs[order(doc_size, decreasing = TRUE)[1]]
sotu[["text"]][idx]

```

Another functionality included in the ```mallet``` R package is to (hierarchically) cluster the topics to assess what topics that are "closer" to each other. Use ```?mallet.topic.hclust``` to see further details on how to cluster topics.

```{r, fig.height=5, fig.width=5}
topic_labels <- mallet.topic.labels(topic.model, num.top.words = 2)
topic_clusters <- mallet.topic.hclust(doc.topics, topic.words, balance = 0.5)
plot(topic_clusters, labels=topic_labels, xlab = "", )
```

## Save and load topic states

We can also store our current topic model state to use it for postprocessing. We can store the state file either as a text file or a compressed gzip file.

```{r}
state_file <- file.path(tempdir(), "temp_mallet_state.gz")
save.mallet.state(topic.model = topic.model, state.file = state_file)
```

We also store the topic counts per document and remove the old model.

```{r}
doc.topics.counts <- mallet.doc.topics(topic.model, smoothed=FALSE, normalized=FALSE)

rm(topic.model)
```

To initialize a model with the sampled topic indicators, one needs to create a new model, load the same data and then load the topic indicators. Unfortunately, setting the alpha parameter vector is currently not possible, so it is not currently possible to initialize the model with the same alpha prior.

```{r}
new.topic.model <- MalletLDA(num.topics=10, alpha.sum = 1, beta = 0.1)
new.topic.model$loadDocuments(sotu.instances)
load.mallet.state(topic.model = new.topic.model, state.file = state_file)

doc.topics.counts[1:3, 1:6]
mallet.doc.topics(new.topic.model, smoothed=FALSE, normalized=FALSE)[1:3, 1:6]
```

This vignette gives a first example of using the mallet R package for topic modelling.

## Save and load topic models

We can also save Mallet topic models and load them back into R.

```{r}
model_file <- file.path(tempdir(), "temp_mallet.model")
mallet.topic.model.save(new.topic.model, model_file)
read.topic.model <- mallet.topic.model.read(model_file)

doc.topics.counts[1:3, 1:6]
mallet.doc.topics(read.topic.model, smoothed=FALSE, normalized=FALSE)[1:3, 1:6]
```

