---
title: "Introduction to R mallet"
author: "David Mimno and Måns Magnusson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mallet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installation

To use the mallet R package, we need to use `rJava`, an R package for using Java within R (to access the mallet Java code). See details at [github.com/s-u/rJava](https://github.com/s-u/rJava).

Next, install the ```mallet``` R package from CRAN. To install, simply use ```install.packages()```

```{r, eval=FALSE}
install.packages("mallet")
```

## Usage

Depending on the size of your data, it can be so that you need to increase the Java virtual machine (JVM) heap memory to handle larger corpora. To do this, you need to specify how much memory you want to allocate to the JVM using the ```Xmx``` flag. Below is an example of allocating 4 Gb to the JVM. 

```{r, eval=FALSE}
options(java.parameters = "-Xmx4g")
```

To load the package, use ```library()```.

```{r}
library(mallet)
```

We start by using the example data set of the State of the Union addresses from 1946 to 2000.

```{r}
library(dplyr)
data(sotu)
sotu[["text"]][1:2]
```

Mallet comes with five different stop list files.

```{r}
mallet_supported_stoplists()
stopwords_en_file_path <- mallet_stoplist_file_path("en")
```

As a first step, we need to create an LDA trainer object and supply the trainer with documents. We start by creating a mallet instance list object. 

This function has a few extra options (whether to lowercase or how we define a token). See ```?mallet.import``` for details.

```{r}
sotu.instances <- 
  mallet.import(id.array = row.names(sotu), 
                text.array = sotu[["text"]], 
                stoplist = stopwords_en_file_path,
                token.regexp = "\\p{L}[\\p{L}\\p{P}]+\\p{L}")
```

If the data is already cleaned and we want to use the index of `text.array`, we can just supply the `text.array`.

```{r}
sotu.instances.short <- 
  mallet.import(text.array = sotu[["text"]])
```

It is also possible to supply stop words as a character vector. 

```{r}
stop_vector <- readLines(stopwords_en_file_path)
sotu.instances.short <- 
  mallet.import(text.array = sotu[["text"]], 
                stoplist = stop_vector)
```


We first need to create a topic trainer object to fit a model.

```{r}
topic.model <- MalletLDA(num.topics=10, alpha.sum = 1, beta = 0.1)
```

Load our documents. We could also pass in the filename of a saved instance list file we build from the command-line tools.

```{r}
topic.model$loadDocuments(sotu.instances)
```

To get the model's vocabulary, we use the method `getVocabulary()`. The vocabulary may be helpful in further curating the stopword list.

```{r}
vocabulary <- topic.model$getVocabulary()
head(vocabulary)
```

Optimize hyperparameters (\code{alpha} and \code{beta}) every 20 iterations, after 50 burn-in iterations.

```{r}
topic.model$setAlphaOptimization(20, 50)
```

Now train a model. Note that hyperparameter optimization is on by default. We can specify the number of iterations. Here we'll use a large-ish round number.

```{r}
topic.model$train(200)
```

We can also run through a few iterations where we pick the best topic for each token rather than sampling from the posterior distribution.

```{r}
topic.model$maximize(10)
```

Get the probability of topics in documents and the probability of words in topics. By default, these functions return raw word counts. Here we want probabilities, so we normalize and add "smoothing" so that nothing has exactly 0 probability.

```{r}
doc.topics <- mallet.doc.topics(topic.model, smoothed=TRUE, normalized=TRUE)
topic.words <- mallet.topic.words(topic.model, smoothed=TRUE, normalized=TRUE)
```

What are the top words in topic 2? Notice that R indexes from 1 and Java from 0, so this will be the topic that mallet called topic 1.

```{r}
mallet.top.words(topic.model, word.weights = topic.words[2,], num.top.words = 5)
```

Show the largest document with at least 50% tokens belonging to topic 2. Note, since the model is not identified, you might end up with another topic if you run the same code. 

```{r}
docs <- which(doc.topics[,2] > 0.50)
doc_size <- nchar(sotu[["text"]])[docs]
idx <- docs[order(doc_size, decreasing = TRUE)[1]]
sotu[["text"]][idx]
```


## Save and load topic states

We can also store our current topic model state to use it for postprocessing. We can store the state file either as a txt file or a compressed gzip file.

```{r}
state_file <- file.path(tempdir(), "temp_mallet_state.gz")
save.mallet.state(topic.model = topic.model, state.file = state_file)
```

We also store the topic counts per document and remove the old model.

```{r}
doc.topics.counts <- mallet.doc.topics(topic.model, smoothed=FALSE, normalized=FALSE)

rm(topic.model)
```

To initialize a model with the sampled topic indicators, one needs to create a new model, load the same data and then load the topic indicators. Unfortunately, it is not possible to set the alpha parameter vector, so it is not currently possible to initialize the model with the same alpha prior.

```{r}
new.topic.model <- MalletLDA(num.topics=10, alpha.sum = 1, beta = 0.1)
new.topic.model$loadDocuments(sotu.instances)
load.mallet.state(topic.model = new.topic.model, state.file = state_file)

doc.topics.counts[1:3, 1:6]
mallet.doc.topics(new.topic.model, smoothed=FALSE, normalized=FALSE)[1:3, 1:6]
```

This vignette gives a first example of using the mallet R package for topic modelling.

## Save and load topic models

We can also save Mallet topic models and load them back into R.

```{r}
model_file <- file.path(tempdir(), "temp_mallet.model")
mallet.topic.model.save(new.topic.model, model_file)
read.topic.model <- mallet.topic.model.read(model_file)

doc.topics.counts[1:3, 1:6]
mallet.doc.topics(read.topic.model, smoothed=FALSE, normalized=FALSE)[1:3, 1:6]
```
