---
title: "Introduction to R mallet"
author: "David Mimno and Måns Magnusson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mallet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Installation

To use the mallet R package, we need to use `rJava`, an R package for using Java within R (to access the mallet Java code). See details at [github.com/s-u/rJava](https://github.com/s-u/rJava).

Next, install the ```mallet``` R package from CRAN. To install, simply use ```install.packages()```

```{r, eval=FALSE}
install.packages("mallet")
```

## Usage

Depending of the size of your data it can be so that you need increase the Java virtual machine (JVM) heap memory to handle larger corpora. To do this you need specify how much memory you want to allocate to the JVM using the ```Xmx``` flag. Below is an example of allocating 4 gB to the JVM. 

```{r, eval=FALSE}
options(java.parameters = "-Xmx4g")
```

To load the package, simply use ```library()```.

```{r}
library(mallet)
```



We start out by using the example data set of the State of the Union adresses from 1790 to 2009.

```{r}
library(dplyr)
data(sotu)
sotu[["text"]][1:2]
```

Mallet comes with five different stop list files.

```{r}
dir(system.file("stoplists/", package = "mallet"))
stopwords_en <- system.file("stoplists/en.txt", package = "mallet")
```

As a first step we need to create a LDA trainer object and supply the trainer with documents. We start out by creating a mallet instance list object. 

This function has a few extra options (whether to lowercase, how we define a token). See ```?mallet.import``` for details.

```{r}
sotu.instances <- 
  mallet.import(id.array = row.names(sotu), 
                text.array = sotu[["text"]], 
                stoplist = stopwords_en,
                token.regexp = "\\p{L}[\\p{L}\\p{P}]+\\p{L}")
```

If the data is already cleaned and we just want to use the index of `text.array` we can simply just supply the `text.array`.

```{r}
sotu.instances.short <- 
  mallet.import(text.array = sotu[["text"]])
```

It is also possible to supply stop words as a character vector. 

```{r}
stop_vector <- readLines(stopwords_en)
sotu.instances.short <- 
  mallet.import(text.array = sotu[["text"]], 
                stoplist = stop_vector)
```


To fit a model we first need to create a topic trainer object.

```{r}
topic.model <- MalletLDA(num.topics=10, alpha.sum = 1, beta = 0.1)
```

Load our documents. We could also pass in the filename of a saved instance list file that we build from the command-line tools.

```{r}
topic.model$loadDocuments(sotu.instances)
```

Get the vocabulary, and some statistics about word frequencies. These may be useful in further curating the stopword list.

```{r}
vocabulary <- topic.model$getVocabulary()
head(vocabulary)

word.freqs <- mallet.word.freqs(topic.model)
head(word.freqs)
```

Optimize hyperparameters (\code{alpha} and \code{beta}) every 20 iterations, after 50 burn-in iterations.

```{r}
topic.model$setAlphaOptimization(20, 50)
```

Now train a model. Note that hyperparameter optimization is on, by default. We can specify the number of iterations. Here we'll use a large-ish round number.

```{r}
topic.model$train(200)
```

We can also run through a few iterations where we pick the best topic for each token, rather than sampling from the posterior distribution.

```{r}
topic.model$maximize(10)
```

Get the probability of topics in documents and the probability of words in topics. By default, these functions return raw word counts. Here we want probabilities, so we normalize, and add "smoothing" so that nothing has exactly 0 probability.

```{r}
doc.topics <- mallet.doc.topics(topic.model, smoothed=TRUE, normalized=TRUE)
topic.words <- mallet.topic.words(topic.model, smoothed=TRUE, normalized=TRUE)
```

What are the top words in topic 2? Notice that R indexes from 1 and Java from 0, so this will be the topic that mallet called topic 1.

```{r}
mallet.top.words(topic.model, word.weights = topic.words[2,], num.top.words = 5)
```

Show the first document with at least 5% tokens belonging to topic 2.

```{r}
sotu[["text"]][doc.topics[,2] > 0.05][1]
```


## Save and load topic states

We can also store our current topic model state to use it for postprocessing. We can store the state file either as a txt file or a compressed gzip file.

```{r}
state_file <- file.path(tempdir(), "temp_mallet_state.gz")
save.mallet.state(topic.model = topic.model, state.file = state_file)
```

We also store the topic counts per document and then remove the old model.

```{r}
doc.topics.counts <- mallet.doc.topics(topic.model, smoothed=FALSE, normalized=FALSE)

rm(topic.model)
```

To initialize a model with the sampled topic indicators one need to create a new model, load the same data and then load the topic indicators. Unfortunately it is not possible to set the alpha parameter vector so it is not currently possible to initialize the model with the exact same alpha prior.

```{r}
new.topic.model <- MalletLDA(num.topics=10, alpha.sum = 1, beta = 0.1)
new.topic.model$loadDocuments(sotu.instances)
load.mallet.state(topic.model = new.topic.model, state.file = state_file)

doc.topics.counts[1:3, 1:10]
mallet.doc.topics(new.topic.model, smoothed=FALSE, normalized=FALSE)[1:3, 1:10]
```

This just gives a first example of how to use the mallet R package for topic modeling.
